---
title: "Untitled"
author: "Lidia Ruiz"
date: "Thursday, May 21, 2015"
output: html_document
---
# Introduction

"Using devices such as Jawbone Up, Nike FuelBand, and Fitbit it is now possible to collect a large amount of data about personal activity relatively inexpensively. These type of devices are part of the quantified self movement - a group of enthusiasts who take measurements about themselves regularly to improve their health, to find patterns in their behavior, or because they are tech geeks. One thing that people regularly do is quantify how much of a particular activity they do, but they rarely quantify how well they do it. In this project, your goal will be to use data from accelerometers on the belt, forearm, arm, and dumbell of 6 participants. They were asked to perform barbell lifts correctly and incorrectly in 5 different ways. More information is available from the website here: http://groupware.les.inf.puc-rio.br/har (see the section on the Weight Lifting Exercise Dataset). "

We will use the datasets available in:

https://d396qusza40orc.cloudfront.net/predmachlearn/pml-training.csv

https://d396qusza40orc.cloudfront.net/predmachlearn/pml-testing.csv

and build a machine-learning algorithm to predict in which manner did the subjects exercise.

# Data Loading and Preprocessing

The first step is loading the required packages for the project.

```{r results="hide"}
library(caret)
library(e1071)
library(randomForest)
```

Now we download the training and testing datasets:

```{r}
train <- "http://d396qusza40orc.cloudfront.net/predmachlearn/pml-training.csv"

test <- "http://d396qusza40orc.cloudfront.net/predmachlearn/pml-testing.csv"

trainingDataset <- read.csv(url(train), na.strings=c("NA","#DIV/0!",""))
testingDataset <- read.csv(url(test), na.strings=c("NA","#DIV/0!",""))

```

Finally we proccess the training dataset to eliminate those variables with a high number of NAs and those variables with low variance. This procedure is done to reduce the size of the dataset and ease the computation. We also eliminate the 'X' column that represents the row number, the user name, and the timestamp related variables.

```{r}
eliminateColumns <- grep("X|user_name|cvtd_timestamp",names(trainingDataset))
trainingDataset <- trainingDataset[, -eliminateColumns]
nearZero <- nearZeroVar(trainingDataset)
trainingDataset <- trainingDataset[, -nearZero]

NAs <- apply(trainingDataset, 2, function(x) {
    sum(is.na(x))
})
trainingDataset <- trainingDataset[, which(NAs == 0)]

dim(trainingDataset)
```


Given that the cleaned training set is large (56 variables and 19622 observations) the required time to build the prediction model is going to be large. To speed up the computation we are going to take the 80% of the data and create a smaller training data set:

```{r}
training <- createDataPartition(y = trainingDataset$classe, p = 0.2, list = FALSE)
subTraining <- trainingDataset[training,] #training dataset
subTesting <- trainingDataset[-training,] #testing dataset for cross-validation
dim(subTraining)
```

The final training dataset has 56 variables and 3927 observations and is significantly smaller then the previous training dataset. 

Once we have prepared a training dataset with proper dimensions for the computation, we build the prediction models.

# Prediction Models

We choose the Random Forest method to choose the appropriate model due to its high accuracy when it works with a huge number of predictor variables.

```{r}
ctrl   <- trainControl(method = "cv", number = 4, allowParallel = TRUE)
modFit <- train(classe ~ ., method = "rf", data = subTraining, prof = TRUE, trControl = ctrl)
```

## Cross-validation and out of sample error
We perform a cross-validation of the model and quantify the out of sample error.
```{r}
predictions <- predict(modFit, subTesting)
confusionMatrix(predictions, subTesting$classe)
``` 

The model fits the subTesting dataset with a 99.24% of accuracy. The out of sample error is 0.76%. The 95% CI is (0.9909, 0.9937).

```{r}
colnames <- names(trainingDataset)
testingColnames <- colnames[!colnames %in% "classe"]
##Subset the test data by the colnamesInTest
testingDataset <- testingDataset[testingColnames]
##Predict the classe variable for each of the 20 test data.

testingDataset$magnet_dumbbell_z <- as.numeric(testingDataset$magnet_dumbbell_z)
testingDataset$magnet_forearm_y <- as.numeric(testingDataset$magnet_forearm_y)
testingDataset$magnet_forearm_z <- as.numeric(testingDataset$magnet_forearm_z)

```

We perform a cross-validation of the model and quantify the out of sample error.
```{r}
predictions <- predict(modFit, testingDataset)

predictions
``` 
```{r echo=FALSE, result="hide"}
  pml_write_files = function(x){
  n = length(x)
  for(i in 1:n){
    filename = paste0("problem_id_",i,".txt")
    write.table(x[i],file=filename,quote=FALSE,row.names=FALSE,col.names=FALSE)
  }
}

pml_write_files(predictions)
```